The curious world of constructive mathematics

### 2022-09-03 algar-lecture-notes.tex
More exercises

{
\renewcommand{\newpage}{\ }
\renewcommand{\vfill}{\ }
\renewcommand{\vfil}{\ }
{\centering\includegraphics[width=0.3\textwidth]{phantoms}\par}
\maketitle
}

@@

Of the many axioms and inference rules of classical logic, exactly one is
responsible for enabling unconstructive arguments, namely the \emph{principle
of excluded middle}:
\[ \varphi \ \vee\ \neg\varphi. \]
The first proof above used this principle in its very first step. In
constructive mathematics, we abstain from this principle; we build
constructive mathematics on \emph{intuitionistic logic}, which contains neither
this principle nor the (equivalent) \emph{principle of double negation
elimination} stating~$\neg\neg\varphi \Rightarrow \varphi$, and insofar as we
layer a set theory on top of our logical foundation, we abstain from the axiom
of choice (which in presence of other common set-theoretical axioms implies the
principle of excluded middle, see Exercise~\ref{ex:diaconescu}).

@@

In contrast, arguments of the second kind are fine from an intuitionistic point
of view: They are not proper proofs by contradiction, but proofs of negated
statements. That such proofs are valid follows directly from the definition of
negation (which texts on classical logic often also adopt) as a certain
implication:
\[ \neg\psi \defequiv (\psi \Rightarrow \bot), \]
where~``$\bot$'', pronounced ``bottom'', is \emph{absurdity}, a canonical false
statement. (Informally, absurdity is also written as~``$\lightning$'' or~``$1 = 0$''.)
Hence, to establish~$\neg\psi$, we may give a proof of~$\bot$ under
the assumption~$\psi$, just as we may give a proof of~$\beta$ under the
assumption~$\alpha$ if we want to prove~$\alpha \Rightarrow \beta$.

The two proofs of the following fact from number theory demonstrate the
difference:
\begin{prop}\label{prop:sqrt2}The number~$\sqrt{2}$ is not rational.\end{prop}
\begin{proof}[Proof (only valid classically)]
Assume that the claim is false, that is, that the number~$\sqrt{2}$ is
\emph{not not} rational. Then~$\sqrt{2}$ is rational. Hence there are
integers~$a$ and~$b$ with~$\sqrt{2} = a / b$. Thus~$2b^2 = a^2$. This identity
contradicts the uniqueness of the prime factor decomposition, since the
factor~$2$ occurs an odd number of times on the left, but an even number of
times on the right.
\end{proof}
\begin{proof}[Proof (also valid intuitionistically)]
Assume that the number~$\sqrt{2}$ is rational. Then there are integers~$a$
and~$b$, \ldots, a contradiction.
[The theorem on the uniqueness of the prime factor decomposition admits an
intuitionistic proof.]
\end{proof}

@@

This sense of bewilderment is resolved by observing that \emph{even though
constructive mathematicians use the same logical symbols, they have a slightly
different meaning in mind.} When a constructive mathematician argues for some
statement~$\varphi$, they mean that they have an \emph{explicit witness}
for~$\varphi$. This shift in meaning from the classical interpretation is
elaborated by the \emph{Brouwer--Heyting--Kolmogorov interpretation} sketched
below. It is informal and philosophical in nature, but still a useful guide to
the constructive meaning of mathematical statements.

@@

The notion of witnesses is compositional in nature and begins with witnesses for \emph{atomic
statements}, those not built from substatements using the logical
connectives~$\wedge,\vee,\Rightarrow$ or the quantifiers~$\forall,\exists$. For
instance, in formal number theory, the atomic statements are of the form
``$s = t$'',
where~$s$ and~$t$ are terms for natural numbers; statements of this form are so
simple that we can check them directly so that witnesses don't need to supply
particular additional information. We hence decree that true atomic statements
have trivial witnesses and that false atomic statements have no witnesses at
all.

Table~\ref{table:bhk} explains what witnesses of compound statements are. For
instance, a witness of a statement of the form
\[ \forall n \? \NN\_ \bigl(\varphi(n) \Rightarrow \psi(n)\bigr) \]
is a rule explaining how, for any natural number~$n : \NN$, witnesses
for~$\varphi(n)$ give rise to witnesses for~$\psi(n)$.

@@

\begin{exercise}[Constructive status of classical tautologies]
Which of the following classical tautologies can reasonably be expected to
admit constructive proofs?
\begin{enumerate}
\item $\neg(\alpha \vee \beta) \ \Longrightarrow\ \neg\alpha \wedge \neg\beta$
\item $\neg(\alpha \wedge \beta) \ \Longrightarrow\ \neg\alpha \vee \neg\beta$
\item $(\alpha \vee \beta) \wedge \neg\alpha \ \Longrightarrow\ \beta$
\item $\forall M\?P(X)\_ (\exists x\?X\_ x \in M) \vee M = \emptyset$
(already interesting for~$X = \{\star\}$)
\item $\forall n\?\NN\_ (n = 0 \vee n \neq 0)$
\item $\forall x\?\RR\_ (x = 0 \vee x \neq 0)$
\item $\forall x\?\RR\_ (\neg(\exists y\?\RR\_ xy=1) \Rightarrow x = 0)$
\item $\forall z\?\overline{\QQ}\_ (z = 0 \vee z \neq 0)$
\item $\forall f \? \NN \to \{0,1\}\_ (\neg\neg\exists n \? \NN\_ f(n) = 0)
\Rightarrow (\exists n \? \NN\_ f(n) = 0)$ (Markov's principle)
\item $\forall f \? \NN \to \{0,1\}\_ \exists n \? \NN\_ (f(n) = 1
\Rightarrow (\forall m \? \NN\_ f(m) = 1))$ (Drinker's paradox)
\item $\forall f \? \NN \to \{0,1\}\_ (\exists n \? \NN\_ f(n) = 0) \vee
(\forall n \? \NN\_ f(n) = 1)$
\item $\forall f \? \NN_\infty \to \{0,1\}\_ (\exists n \? \NN_\infty\_ f(n) = 0) \vee
(\forall n \? \NN_\infty\_ f(n) = 1)$

@@

\begin{exercise}[Basics on negation]
Recalling that negation is defined as implying absurdity,~$\neg\varphi
\defequiv (\varphi \Rightarrow \bot)$, verify intuitionistically without
recourse to truth tables:
\begin{enumerate}
\item $\varphi \Rightarrow \neg\neg\varphi$
\item $\neg\neg\neg\varphi \Leftrightarrow \neg\varphi$
\item $\neg\neg(\varphi \vee \neg\varphi)$
\item $(\forall\psi\_ (\psi \vee \neg\psi))
\Longleftrightarrow (\forall\psi\_ ((\neg\neg\psi) \Rightarrow
\psi))$

@@

\begin{exercise}[Diaconescu's theorem]
\label{ex:diaconescu}
The axiom of choice can be put as: ``Every surjective map has a
section.'' (A \emph{section}~$s$ to a surjective map~$f$ is a map in the other direction such that~$f \circ s =
\mathrm{id}$.) It is a theorem due to Diaconescu that
the axiom of choice implies the law of excluded middle. To this end, let~$\varphi$ be a
statement and consider the subsets
\begin{align*}
  U &= \{ x \in X \,|\, (x = 0) \vee \varphi \} \\
  V &= \{ x \in X \,|\, (x = 1) \vee \varphi \}
\end{align*}
of the discrete set~$X \defeq \{ 0,1 \}$.
\begin{enumerate}
\item Verify that~$U = V$ if and only if~$\varphi$.
\item Using that~$x = y \vee x \neq y$ for all elements~$x,y \in X$, show that
the existence of a section of the surjective map
\[ \begin{array}{@{}rcl@{}}
  X &\longrightarrow& \{U,V\} \\
  0 &\longmapsto& U \\
  1 &\longmapsto& V
\end{array} \]
implies~$\varphi \vee \neg\varphi$.
\end{enumerate}
\end{exercise}

@@

@@

@@

>>>>

{
\renewcommand{\newpage}{\ }
\renewcommand{\vfill}{\ }
\renewcommand{\vfil}{\ }
{\vspace*{-6em}\centering\includegraphics[width=0.3\textwidth]{phantoms}\par}
\thispagestyle{empty}\enlargethispage{7em}
\maketitle
}

@@

Of the many axioms and inference rules of classical logic, exactly one is
responsible for enabling unconstructive arguments, namely the \emph{principle
of excluded middle}:
\[ \varphi \ \vee\ \neg\varphi. \]
The first proof above used this principle in its very first step. In
constructive mathematics, we abstain from this principle; we build
constructive mathematics on \emph{intuitionistic logic}, which contains neither
this principle nor the (equivalent) \emph{principle of double negation
elimination} stating~$\neg\neg\varphi \Rightarrow \varphi$, and insofar as we
layer a set theory on top of our logical foundation, we abstain from the axiom
of choice (which in presence of other common set-theoretical axioms implies the
principle of excluded middle, see Exercise~\ref{ex:diaconescu}). As a
consequence, we cannot generally reason by contradiction in constructive mathematics.

@@

In contrast, arguments of the second kind are fine from an intuitionistic point
of view: They do not constitute proper proofs by contradiction, but instead are proofs of negated
statements. That such proofs are valid follows directly from the definition of
negation as a certain
implication (which, incidentally, texts on classical logic often also adopt):
\[ \neg\psi \defequiv (\psi \Rightarrow \bot), \]
where~``$\bot$'', pronounced ``bottom'', is \emph{absurdity}, a canonical false
statement. (Informally, absurdity is also written as~``$\lightning$'' or~``$1 = 0$''.)
Hence, to establish~$\neg\psi$, we may give a proof of~$\bot$ under
the assumption~$\psi$, just as we may give a proof of~$\beta$ under the
assumption~$\alpha$ if we want to prove~$\alpha \Rightarrow \beta$.

The two proofs below of the following fact from number theory demonstrate the
difference:
\begin{prop}\label{prop:sqrt2}The number~$\sqrt{2}$ is not rational.\end{prop}
\begin{proof}[Proof (only valid classically)]
Assume that the claim is false, that is, that the number~$\sqrt{2}$ is
\emph{not not} rational. Then~$\sqrt{2}$ is rational. Hence there are
integers~$a$ and~$b$ with~$\sqrt{2} = a / b$. Thus~$2b^2 = a^2$. This identity
contradicts the uniqueness of the prime factor decomposition, since the
factor~$2$ occurs an odd number of times on the left but an even number of
times on the right.
\end{proof}
\begin{proof}[Proof (also valid intuitionistically)]
Assume that the number~$\sqrt{2}$ is rational. Then there are integers~$a$
and~$b$, \ldots, a contradiction.
[The requisite theorem on the uniqueness of the prime factor decomposition admits an
intuitionistic proof.]
\end{proof}

@@

This sense of bewilderment is resolved by observing that \emph{even though
constructive mathematicians use the same logical symbols, they have a slightly
different meaning in mind.} When a constructive mathematician argues for some
statement~$\varphi$, they mean that they have an \emph{explicit witness}
for~$\varphi$. This shift in meaning from the classical interpretation is
elaborated by the \emph{Brouwer--Heyting--Kolmogorov interpretation} sketched
below. It is informal and philosophical in nature and not without
issues~\cite{artemov:bhk,sanz-piecha:critical-bhk,dalen:bhk}, but still a useful guide to
the constructive meaning of mathematical statements.

@@

The notion of witnesses is compositional in nature and starts out with witnesses for \emph{atomic
statements}, those not built from substatements using the logical
connectives~$\wedge,\vee,\Rightarrow$ or the quantifiers~$\forall,\exists$. For
instance, in formal number theory, the atomic statements are of the form
``$s = t$'',
where~$s$ and~$t$ are terms for natural numbers; statements of this form are so
simple that we can check them directly so that witnesses don't need to supply
particular additional information. We hence decree that true atomic statements
have trivial witnesses and that false atomic statements have no witnesses at
all.

Table~\ref{table:bhk} explains what witnesses of compound statements are. For
instance, a witness of a statement of the form
\[ \forall n \? \NN\_ \bigl(\varphi(n) \Rightarrow \psi(n)\bigr) \]
is a rule explaining how, for every natural number~$n : \NN$, witnesses
for~$\varphi(n)$ give rise to witnesses for~$\psi(n)$.

@@

\begin{exercise}[Constructive status of classical tautologies]\label{ex:tautologies}
Which of the following classical tautologies can reasonably be expected to
admit constructive proofs?
\begin{enumerate}
\item $\neg(\alpha \vee \beta) \ \Longrightarrow\ \neg\alpha \wedge \neg\beta$
\item $\neg(\alpha \wedge \beta) \ \Longrightarrow\ \neg\alpha \vee \neg\beta$
\item $(\alpha \Rightarrow \beta) \ \Longrightarrow\ (\neg\alpha \vee \beta)$
\item $(\alpha \vee \beta) \wedge \neg\alpha \ \Longrightarrow\ \beta$
\item $\forall M\?P(X)\_ (\exists x\?X\_ x \in M) \vee M = \emptyset$
(already interesting for~$X = \{\star\}$)
\item $\forall n\?\NN\_ (n = 0 \vee n \neq 0)$
\item $\forall x\?\RR\_ (x = 0 \vee x \neq 0)$
\item $\forall x\?\RR\_ (\neg(\exists y\?\RR\_ xy=1) \Rightarrow x = 0)$
\item $\forall z\?\overline{\QQ}\_ (z = 0 \vee z \neq 0)$
\item $\forall f \? \NN \to \{0,1\}\_ (\neg\neg\exists n \? \NN\_ f(n) = 0)
\Rightarrow (\exists n \? \NN\_ f(n) = 0)$ (Markov's principle)
\item $\forall f \? \NN \to \{0,1\}\_ \exists n \? \NN\_ (f(n) = 1
\Rightarrow (\forall m \? \NN\_ f(m) = 1))$ (Drinker's paradox)
\item $\forall f \? \NN \to \{0,1\}\_ (\exists n \? \NN\_ f(n) = 0) \vee
(\forall n \? \NN\_ f(n) = 1)$
\item $\forall f \? \NN_\infty \to \{0,1\}\_ (\exists n \? \NN_\infty\_ f(n) = 0) \vee
(\forall n \? \NN_\infty\_ f(n) = 1)$

@@

\begin{exercise}[Basics on negation]
Recalling that negation is defined as implying absurdity,~$\neg\varphi
\defequiv (\varphi \Rightarrow \bot)$, verify intuitionistically without
recourse to truth tables:
\begin{enumerate}
\item $\varphi \Rightarrow \neg\neg\varphi$
\item $\neg\neg\neg\varphi \Leftrightarrow \neg\varphi$
\item $\neg\neg(\varphi \vee \neg\varphi)$
\item $\neg\neg(\alpha \wedge \beta) \Leftrightarrow (\neg\neg\alpha \wedge
\neg\neg\beta)$
\item $(\neg\neg\alpha \wedge (\alpha \Rightarrow \neg\neg\beta)) \Longrightarrow
\neg\neg\beta$
\item $(\forall\psi\_ (\psi \vee \neg\psi))
\Longleftrightarrow (\forall\psi\_ ((\neg\neg\psi) \Rightarrow
\psi))$

@@

\begin{exercise}[Brouwerian counterexamples]
Show that each of the following statements implies the law of excluded middle,
hence is not available in constructive mathematics.
\begin{enumerate}
\item Every ideal of~$\ZZ$ is finitely generated.

{\noindent\scriptsize\emph{Hint.} Use that finitely generated ideals of~$\ZZ$
are principal ideals and consider the ideal~$\aaa \defeq \{ x \in \ZZ
\,|\, x = 0 \vee \varphi \}$.\par}

{\noindent\scriptsize\emph{Remark.} The failure of every ideal of~$\ZZ$ to be
finitely generated should not be misconstrued to exclaim that in constructive
mathematics, there suddenly would be ideals of~$\ZZ$ of infinite rank. The
failure is simply because, given an abstract ideal, we cannot pinpoint a finite
system of generators.\par}

\item Over every field, the polynomial~$X^2 + 1$ is either reducible or
irreducible.

{\noindent\scriptsize\emph{Hint.} Consider the field~$K \defeq \{ z \in \QQ(i)
\,|\, z \in \QQ \vee \varphi \}$.\par}

\item Subsets of Kuratowski-finite sets are Kuratowski-finite.

{\noindent\scriptsize\emph{Note.} A set~$X$ is \emph{Kuratowski-finite} if and
only if, for some number~$n \in \NN$, there is a surjective map~$[n] \to X$,
where~$[n] = \{ 0,1,\ldots,n-1 \}$. More briefly, a set~$X$ is
Kuratowski-finite iff its elements can be enumerated: $X = \{x_1,\ldots,x_n \}$.\par}
\end{enumerate}
Show that the following statement implies Markov's principle (from
Exercise~\ref{ex:tautologies}):
\begin{enumerate}
\addtocounter{enumi}{3}
\item Every real number which is not zero is invertible.
\end{enumerate}
\noindent
More Brouwerian counterexamples can be
found in~\cite{mandelkern:brouwerian-counterexamples} and
in~\cite{mines-richman-ruitenburg:constructive-algebra}.
\end{exercise}

\begin{exercise}[Diaconescu's theorem]
\label{ex:diaconescu}
The axiom of choice can be put as: ``Every surjective map has a
section.'' (A \emph{section}~$s$ to a surjective map~$f$ is a map in the other direction such that~$f \circ s =
\mathrm{id}$.) A theorem of Diaconescu states that
the axiom of choice implies the law of excluded middle. To this end, let~$\varphi$ be a
statement and consider the subsets
\begin{align*}
  U &= \{ x \in X \,|\, (x = 0) \vee \varphi \} \\
  V &= \{ x \in X \,|\, (x = 1) \vee \varphi \}
\end{align*}
of the discrete set~$X \defeq \{ 0,1 \}$.
\begin{enumerate}
\item Verify that~$U = V$ if and only if~$\varphi$.
\item Using that~$x = y \vee x \neq y$ for all elements~$x,y \in X$, show that
the existence of a section of the surjective map
\[ \begin{array}{@{}rcl@{}}
  X &\longrightarrow& \{U,V\} \\
  0 &\longmapsto& U \\
  1 &\longmapsto& V
\end{array} \]
implies~$\varphi \vee \neg\varphi$.
\end{enumerate}
\end{exercise}

@@

\chapter{Toposes and their internal language}

\section{Examples}
\newpage

\section{Definition}
\newpage

\section{The Kripke--Joyal semantics of sheaf toposes}
\newpage

@@

\section{A fantastical ring}
\newpage

\section{The generic prime filter}
\newpage

\section{Explicit construction}
\newpage

@@

Cite surveys of .bib
